% !TEX root = thesis.tex

There are many people who have contributed to my experience at Princeton. 
%
Firstly, I would like to thank my advisor, Dr. Yannis Kevrekidis, for agreeing to take me as a student and guiding me through my doctoral work. 
%
I am grateful to Dr. Pablo Debenedetti for helpful research discussions and input, especially during the first few years. 
%
I am also extremely grateful to Dr. Stanislav Shvartsman, for providing both interesting research problems as well as helpful life discussions and advice. 
%
I would like to thank all of my collaborators, most notably Dr. Ronen Talmon, Dr. C. Wiliam Gear, and Dr. Bomyi Lim, whose interesting research ideas and perspectives greatly influenced my work and outlook.
%
I am also grateful for the Department of Energy Computational Science Graduate Fellowship (grant number DE-FG02-97ER25308) and the National Science Foundation Graduate Research Fellowship Program (grant number DGE 1148900) for funding my research. 

The Chemical and Biological Engineering Department at Princeton has been a source of great support over the years.
%
I would like to thank Karen for all of her patience and help with navigating the graduate school process. 
%
I am thankful for my lab mates over the years for their help, advice, and interesting conversations.
%
I would also like to thank all of my classmates for being one of the most fun and colorful groups of people I have known; you truly made the graduate school experience one to remember!
%
I would specifically like to thank Kim and Raleigh for fun and interesting lunch time conversations.

I would also like to thank my family, and more specifically my parents for all of their love and support over the years. 
%
From a young age, they instilled in me a work ethic and love of learning which has proven invaluable. 
%
Without them, I most certainly would not be where I am today. 

And lastly, I would like to thank my husband, Matthew, for his continued love and encouragement. 
%
Five years ago, he generously told me that he would follow me wherever I wanted to go to graduate school. 
%
Through the good and bad times, he has been by my side. 
%
He has celebrated with me, listened to me complain, offered me advice, and through it all been the person who I know I can rely on unconditionally. 
